Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-07-25 MIMOSA.tex
Rewrote the conclusion.

\subsection{ICS}
\subsubsection*{HVTN065}

Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples\footnote{Were these stimulations described in the data section}. 
The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders when the magnitude of response was small (i.e. ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs. 0.923 for Fisher's exact test vs. 0.827 for LRT vs. 0.775 for ranked fold change). When the magnitude of response was larger (GAG-1-PTEG stimulation), MIMOSA performed as well as Fisher's exact test, and better than other competing methods (Figure~\ref{fig:HVTN065} C) (AUC=0.982 for MIMOSA vs. 0.978 for Fisher's exact test vs. 0.877 for LRT vs. 0.784 for ranked fold change). Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D). These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}) \footnote{Perhaps we could only keep the Env stimulation}.

\subsubsection*{HVTN054}

\subsection{Single-cell gene expression}
We applied the MIMOSA model to Fluidigm single-cell gene expression data from CD8+ T-cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. 
This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation. In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} A-C show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm}). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions (Figure~\ref{fig:fluidigm} B,C). A similar analysis using a two-sided Fisher's exact test (Figure xx) does not reveal any stimulation specific changes of expression.\footnote{Perhaps it would be good to compare the number of differentially expressed genes}.

@@

\subsection{Parameter estimation}
Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Supplementary Material). For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
\begin{align}
\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
\end{align}
where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by 
\begin{align}
\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
\end{align}
The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger. In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2. It becomes highly dependent on the initial values and can even fail to converge when good initial values are provided. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large. More details about both algorithms are given in supplementary material.

@@

\section{Discussion}
The variety of single-cell assays being adopted by the immunology community is increasing. Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single-cell count data. Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}. Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.

The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model counts in experimental units (i.e. individuals) across multiple conditions (i.e. vaccine responders and non--responders). Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}). The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.

Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}). Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination, the model is naive to vaccine time point. Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C). Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182. None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison. Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).

Although ICS is the motivating example for the MIMOSA model, it can be applied any type of single--cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions. To date, the typical approach to Fluidigm single cell gene expression analysis has been focused on identifying differences in the continuous part of the signal, corresponding to the $C_t$ values and the ``missing'' data from cells that don't express a given gene has generally been ignored, or used as part of ad--hoc pre-filtering\cite{Flatz:2011jb}. The ability of MIMOSA to identify stimulation--specific expression patterns in single--cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C). MIMOSA explicitly models and performs inference on the observed positive cell counts between different conditions. Additionally, extending the model to take into account multiple  conditions is straightforward.

Detecting differences in polyfunctional cell populations (i.e. identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in vaccine research, since it allows the identification of more precisely defined, more homogeneous cell popualtions~\cite{Milush:2009bz}. These polyfunctional cell populations may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design. What's more, changes in these polyfunctional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines, or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e. mixtures of responders and non--responders).  Sometimes these tests are entirely empirical in nature (i.e. thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}. As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}. Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single--cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).

\section{Conclusions}
We have developed a mixture model framework that we call MIMOSA .for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays that frequently arise in the context of vaccine trials and immunological studies. Our model can be applied in a univariate or multivariate fashion, and we have shown that it has greater sensitivity and specificity than other typical approaches used to analyze the type of count--based data derived from single cell assays, including Fisher's exact test, the non-Bayesian likelihood ratio test, or ad--hoc rank based approaches. The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).

%\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
%Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold. Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods. Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines. In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
%
%Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach. We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples). For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively. Note that usually, $N_u \ne N_s$. The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants. For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table: 
% latex table generated in R 2.14.0 by xtable 1.5-6 package
% Wed Oct 12 09:41:50 2011

>>>>

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change\footnote{Why ranked?}, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}) \footnote{Perhaps we could only keep the Env stimulation}.

\subsection{Single-cell gene expression}
We applied the MIMOSA model to Fluidigm single-cell gene expression. This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation. In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} A-C show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm}). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions (Figure~\ref{fig:fluidigm} B,C). A similar analysis using a two-sided Fisher's exact test (Figure xx) does not reveal any stimulation specific changes of expression.\footnote{Perhaps it would be good to compare the number of differentially expressed genes}.

@@

\subsection{Parameter estimation}
Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Supplementary Material). For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
\begin{align}
\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
\end{align}
where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by 
\begin{align}
\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
\end{align}
The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger. In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than three. It becomes highly dependent on the initial values and can fail to converge in many instances. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large. More details about both algorithms are given in Web Supplementary material.

@@

\section{Discussion}
Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few. Single cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single cell). As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}. In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single cell assay.

The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental individuals across multiple conditions (\textit{i.e.} vaccine responders and non-responders). Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}). The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample. This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.

%Importantly, the analysis of real-world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}). Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination, the model is naive to vaccine time point. Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre-vaccination) and 182 (post-vaccination) than the current standard approach (Fisher's exact test), or non-Bayesian alternatives such as the likelihood ratio test or ranking by log-fold change (Figure~\ref{fig:HVTN065} A,C). Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182. None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison. Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).

Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions. In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or simply using this information for pre-filtering \cite{Flatz:2011jb}. The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).

Detecting differences in poly-functional cell populations (\textit{i.e.} identifying changes in cell populations that co-express multiple proteins, cytokines, or genes), is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\cite{Milush:2009bz}. In the context of HIV, these poly-functional cell populations have been shown to be correlated with good outcome such as vaccine protection and long-term disease non-progression (ADD REFERENCES TO THE PAPERS STEVE SUGGESTED IN THE GRANT). In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected. It  follows that if a specific cell subset expressing multiple markers is being differentially expressed, differential expression based on the marginal cell counts should also be detected. As such, identifying poly-functional cytokine profiles from ICS data can be done in an iterative way. First univariate tests on marginal populations are performed and then specific cell subsets expressing the positive markers detected are then tested. Even in this situation, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that needs to be tested, and a multivariate approach might be preferable. In this case, as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\cite{Nason:2006dx}.

When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression. Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B). Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. A solution would be to explore alternative model parametrizations that could be used to reduce the number of required parameters. For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. More exploratory work will need to be done in this area once high dimensional single cell level data with large number of samples become available.

All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).

\section{Supplementary Materials}
Web Appendices, Tables, and Figures (limit to the types of material actually in question) referenced in Sections (give section numbers) are available in the attached Web-based supplementary material.
% This will be replaced at submission time
%at the Biometrics website on Wiley Online Library.
\section{Acknowledgments}
%\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
%Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold. Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods. Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines. In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
%
%Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach. We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples). For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively. Note that usually, $N_u \ne N_s$. The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants. For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table: 
% latex table generated in R 2.14.0 by xtable 1.5-6 package
% Wed Oct 12 09:41:50 2011
